Topological order in the quantum dimer-pentamer model

### 2015-09-09 qdpm_paper.tex
Reformatted gap table.
Worked on introduction. Added placeholder figure 1.

@@

\begin{abstract}

\end{abstract}

@@

    It became apparent in the 80's and early 90's that the Landau theory of symmetry breaking did not
    encompass all possible types of order that can be found in the groundstate of a condensed matter system. 
    In fact it was found that the symmetry of a system
    could increase as a system was cooled past a critical tempature and a phase transition occured
    that did not fit into Landau's theory. In these cases the odering 
    that could not be described with an order
    parameters~\cite{Wen1990} was a topological order. Though the discovery of topological order was originally thought
    to help understand superconductivity in the context of chiral spin systems \cite{HERE} it was
    not able to do so. However, due to the robust ground state degeneracy of systems exhibiting
    topological order they have become candidates for methods of storing and manipulating
    quantum information.  Also important to the field of quantum information, but interesting in
    its own right, is that topologically ordered systems exhibit fractal excitations.

    One of the simplest and most well studied models that exhibits topological order is the
    toric code, an exactly solvable model. Unfortunately it is
    a difficult model to physically realize. Another class of models that are well studied are
    the quantum dimer models (QDM).  These models can exhibit topological order and are also
    easier to experimentally realize, for example some short range resonating valence bond
    systems can be approximated with a QDM that provides a low energy effective description of
    the frustrated spins. Some correlation functions in QDMs are solvable using Pfaffian models
    but in general they must be studied numerically.

    In this paper we numerically study the quantum dimer pentamer model (QDPM) on the square
    lattice at the RK point.  The QDPM contains an exact $Z_3$ local gauge symmetry due to the
    introduction of pentamer objects and we believe it exhibits $Z_3$ topological order.
    Pentamers are objects consisting of four dimers touching a vertex.  The Hilbert space
    comprises all fully packed dimer and pentamer coverings. The presence or absence of a dimer
    may be defined in terms of Ising degrees of freedom that live on the links of the square
    lattice and this distinguishes it from the $Z_3$ toric code with three degrees of freedom on
    the links. We show numerically that a liquid dimer pentamer phase exists at the RK point
    with exponentially decaying dimer correlations. We also estimate the dimer and vison gaps
    showing that the $Z_3$ visons are the lowest lying and most fundamental excitations.

\section{Background}

@@

    \begin{table}[htpb]
        \centering
        \caption{$\Delta \tau = 150$}
        \label{tab:label}
        \begin{tabular}{c c c c}
            \hline\hline
         L            & 8x8 Gap                               & 12x12                            & 16x16                               & 18x18 \\ 
            \hline                                                                                                                           
         frac 1.0     &                                       &                                  &                                     &  \\
            \hline                                                                                                                           
         Dimer        &  $ 0.03230 \pm 0.01142$               & $ 0.01430 \pm 0.00034 $          & $0.00806 \pm 0.00023$               & \\
         $Z_2$ origin &  $ 0.00787 \pm 5.3\times 10^{-6} $    & $ 0.00344 \pm 9.0\times 10^{-7}$ & $0.00198 \pm 3.8\times 10^{-6} $    &  \\
         $Z_3$ origin &  NA                                   & NA                               & NA                                  & \\
            \hline                                                                                                                         
         frac 0.9     &                                       &                                  &                                     &  \\
            \hline                                                                                                                          
         Dimer origin &  $ 0.00419 \pm 8.1\times 10^{-6} $    & $ 0.00406 \pm 1.7\times 10^{-5} $ & $0.00429 \pm 6.6\times 10^{-6} $   &  $ 0.00405 \pm 1.6\times 10^{-5} $\\
         $Z_2$ origin &  $ 0.00216 \pm 1.5\times 10^{-6} $    & $ 0.00141 \pm 8.4\times 10^{-7} $ & $0.00129 \pm 1.8\times 10^{-6} $   &  $ 0.00126 \pm 1.7\times 10^{-6} $\\
         $Z_3$ origin &  $ 0.00346 \pm 3.8\times 10^{-6} $    & $ 0.00348 \pm 5.1\times 10^{-6} $ & $0.00359 \pm 3.5\times 10^{-6} $   &  $ 0.00355 \pm 3.3\times 10^{-6}$\\
            \hline                                                                                                                          
         frac 0.5     &                                       &                                   &                                    & \\
            \hline                                                                                                                          
         Dimer origin &                                       & $ 0.02075 \pm 8.9\times 10^{-5} $ & $0.02049 \pm 0.00011 $             &  $ $ \\
         $Z_2$ origin &                                       & $ 0.00615 \pm 2.2\times 10^{-6} $ & $0.00614 \pm 2.8\times 10^{-6} $   &  $ $ \\
         $Z_3$ origin &                                       & $ 0.01939 \pm 2.2\times 10^{-5} $ & $0.01927 \pm 0.00002 $             &  $ $  
        \end{tabular}
    \end{table}

>>>>

\newcommand{\figref}[1]{Fig. \ref{#1}}
\newcommand{\secref}[1]{Sec. \ref{#1}}

@@

\begin{abstract}
We study the ground state of the quantum dimer-pentamer model (QDPM) on the square lattice. This model is a generalization of the square lattice quantum dimer model (QDM) as its configuration space comprises fully-packed hard-core dimer coverings as well as configurations containing pentamers, where four dimers touch a vertex. Thus in the QDPM, the fully-packed, hard-core constraint of the QDM is relaxed such that the local dimer number at each vertex is fixed modulo 3; correspondingly, the local $U(1)$ gauge symmetry of the QDM Hilbert space is reduced to a local $Z_3$ gauge symmetry in the QDPM. We construct a local Hamiltonian for which the Rokhsar-Kivelson (RK) state (the equal superposition of all configurations in a topological sector) is the exact ground state and has a 9-fold topological degeneracy on the torus. Using Monte Carlo calculations, we find no spontaneous symmetry breaking in the RK wavefunction and that its dimer-dimer correlation function decays exponentially. Additionally, we discuss the possibility of $Z_3$ topological order in the ground state of the QDPM.
\end{abstract}

@@

Over the last several decades, experimental and theoretical work has demonstrated that strongly interacting quantum many-body systems support exotic quantum phases phases of matter not described by the Landau symmetry breaking paradigm. One class of such quantum phases of matter are topologically ordered quantum liquids~\cite{Wen1990}, first realized experimentally via the fractional quantum Hall effect~\cite{FQHE????}. Despite the absence of conventional symmetry breaking and a local order parameter, topological quantum liquids possess a non-trivial quantum order that distinguishes them from trivial liquid phases~\cite{Nayak2008}. Although not detectable by local operators, topological order arises due to a non-local entanglement structure that leaves a signatures in the bipartite entanglement entropy~\cite{Levin2006a,Kitaev2006b}. Such topological phases provide the basis for a proposal for fault-tolerant quantum information processing~\cite{Freedman2001,Kitaev2003}.

Because of the exotic nature of topological phases and their potential use for quantum information processing, there is great interest in the development of theoretical models which display these phases. Such models offer the possibility of furthering the understanding of these phases as well as providing possibly candidate experimental systems in which to constructively engineer topological order~\cite{Duan2003,Jaksch2005,Lewenstein2007,Jiang2008d,Weimer2010a,Herdman2010c,Martinis2015}. A variety of exactly soluble lattice models are known to support ground states with topological order that ranges from the simplest variety, so-called $Z_2$ topological order~\cite{Kitaev2003,Wen2003}, to those with non-Abelian anyonic excitations~\cite{Levin2005a,Kitaev2006a}. These exactly soluble models generally involve complex many-body interactions, and thus it is desirable to find simpler models supporting topological order that are closer to what is accessible experimentally.

Among the simplest class of models displaying topological quantum liquid phases are those with local geometric constraints; the frustration inherent in these constraints inhibits the formation of symmetry breaking local order and given sufficient dynamics, such models may have symmetric liquid ground states. The simplest such geometrically constrained model is the quantum dimer model (QDM) describing dimer degrees of freedom living on the links of a lattice~\cite{Rokhsar1988,Moessner2008}; the hard-core, fully-packed QDM has a local constraint that requires exactly one dimer touching each vertex.

\begin{figure}[b!]
    \centering
    \includegraphics[width=0.5\columnwidth]{QDPM_ex_config.pdf}
    \caption{Example of an allowed configuration in the quantum dimer-pentamer model.}
    \label{fig:QDPMex}
\end{figure}

While on non-bipartite lattices, the QDM posses a topologically ordered ground state~\cite{Moessner2001a,Fendley2002}, on the square lattice there is only a liquid ground state at an isolated critical quantum point~\cite{Leung1996,Syljuasen2006}. If the hard-core constraint of the QDM is relaxed to a parity constraint such that each vertex much have an even (odd) number of dimers touching each vertex, the simplest non-trivial dynamics lead to the (odd) toric code which has a $Z_2$ topologically ordered ground state~\cite{Kitaev2003,Wen2003}.

The existence of topologically ordered ground states in dimer-based models is intimately related to the local constraint which imposes a local gauge symmetry in the low energy Hilbert space; the topological order that arises is directly related to a discrete gauge theory with the appropriate local gauge constraint~\cite{Moessner2001}.

Given that these models with dimer degrees of freedom and local constraints posses the simplest non-trivial local (Ising) Hilbert space at each link, they provide a (possibly) simpler route to engineering exotic phases; the local constraints constraint can often be enforced by a local potential energy penalty and the required dynamics are generally expected to arise at the lowest order non-trivial order in perturbation theory. Previous work as demonstrated that GQDM on the square lattice support gapped $Z_2$ topologically ordered phases as well as gapless quantum critical points. On non-bipartite lattices GQDM supported $Z_2$ topological order~\cite{Moessner2001a,Misguich2002} as well as doubled-semion phases~\cite{Qi2014,Buerschaper2014a}. A full characterization of all exotic phases that can exist within this framework is thus desirable.

In this work we extend this paradigm to a new GQDM on the square lattice, what we term the quantum dimer-pentamer model (QDPM). In the QDPM, the local constraint is relaxed from that of the QDM to require that either $1$ or $4$ dimers to touching each vertex, as shown in figure \figref{fig:QDPMex}. Correspondingly, the QDPM extends the Hilbert space of the QDM to include both fully-packed hard-core dimer configurations, as well as those pentamers, four dimers touch a vertex. To give the system dynamics, the Hamiltonian includes terms that incorporate the simplest non-trivial dynamics that don't violate the local constraint.

We present a numerical study of the ground state of the QDPM at an exactly soluble point and demonstrate the the soluble point represents a gapped, disordered quantum liquid phase.  Using Monte Carlo calculations, we explicitly demonstrate the absence of symmetry breaking order and the exponential decay of correlation functions; by computing imaginary-time correlation functions, we find that the RK point is gapped in the thermodynamic limit. By construction the QDPM has an exact local $Z_3$ symmetry and conserves a $Z_3$ topological winding number. Thus the QDPM has a 9-fold topological degeneracy on the torus. Thus we argue that the ground state of the QDPM posses $Z_3$ topological order.

This this paper is organized as follows. In \secref{sec:Background} we provide relevant background for the QDM and related models. In \secref{sec:QDPM} we present the details of the QPDM. \secref{sec:MC} provides the results of a Monte Carlo study of the ground state of the QDPM at its exactly soluble point.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}
\label{sec:Background}

\subsection{Generalized quantum dimer models}
\label{sec:GQDM}

Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link has either $0$ or $1$ dimers. The Hilbert space consists of configurations of dimers on the square lattice that satisfy a local constraint at each vertex; the simplest constraint is that of the hard-core fully-packed QDM where a single dimer must touch each vertex. We write this constraint as $n_v = 1$, where $n_v$ is the number of dimers that touch a vertex. Other unique, non-trivial constraints are $n_v=2$ and the parity constraints $n_v\mod2 =0$ and  $n_v\mod2 =1$ (see \figref{fig:GQDMexamples}).

The Hamiltonian of a GQDM is of the form
\begin{equation}
H_{\rm{GQDM}} = -t \hat{T} + v \hat{V}
\end{equation}
where $\hat{T}$ is an off-diagonal kinetic term that preserves the constraint, $\hat{V}$ is a local potential energy term, and $v$ and $t$ are the strength of each term.

@@

    \begin{table*}[htpb]
    \setlength\tabcolsep{0.3cm}
        \centering
        \caption{$\Delta \tau = 150$}
        \label{tab:label}
        \begin{tabular}{l l l l l}
            \hline\hline
         L            		& 8x8                               		& 12x12                            		& 16x16					& 18x18 \\ 
            \hline                                                                                                                           
         frac 1.0     	&                                       		&                                  		&                                    		&  \\
            \hline                                                                                                                           
         Dimer        	&  $ 3(1) \times 10^{-2}$        	& $ 1.43(3) \times 10^{-2}$	& $0.81(2)\times 10^{-2}$  	& \\
         $Z_2$ origin 	&  $ 0.787?(5)\times 10^{-2} $ 	& $ 0.344? (1)\times 10^{-2}$ 	& $0.198? (4)\times 10^{-2} $ 	&  \\
         $Z_3$ origin 	&  NA                                   	& NA                               		& NA                                  	& \\
            \hline                                                                                                                         
         frac 0.9     	&                                       		&                                  		&                                     		&  \\
            \hline                                                                                                                          
         Dimer origin 	&  $ 0.419(1)\times 10^{-2} $    	& $ 0.406(2)\times 10^{-2} $ 	& $0.429(1)\times 10^{-2} $   	&  $ 0.405(2)\times 10^{-2} $\\
         $Z_2$ origin 	&  $ 0.216?(2)\times 10^{-2} $ 	& $ 0.141?(1)\times 10^{-2} $ 	& $0.129?(2)\times 10^{-2} $	&  $ 0.126?(2)\times 10^{-2} $\\
         $Z_3$ origin 	&  $ 0.346?(4)\times 10^{-2} $ 	& $ 0.348(1)\times 10^{-2} $ 	& $0.359?(4)\times 10^{-2} $	&  $ 0.355?(3)\times 10^{-2}$\\
            \hline                                                                                                                          
         frac 0.5     	&                                       		&                                   		&                                    		& \\
            \hline                                                                                                                          
         Dimer origin 	&                                       		& $ 2.075(9)\times 10^{-2} $ 	& $2.05(1) \times 10^{-2}$		&  $ $ \\
         $Z_2$ origin 	&                                       		& $ 0.615?(2)\times 10^{-2} $ 	& $0.614?(3)\times 10^{-2} $	&  $ $ \\
         $Z_3$ origin 	&                                       		& $ 1.939(2)\times 10^{-2} $ 	& $1.927(2)\times 10^{-2} $	&  $ $  
        \end{tabular}
    \end{table*}
